# Section 1: Data Loading & Cleaning

### Data Quality Objectives:
1. **Standardize Schema:** Standardize all column names into a consistent `snake_case` format.
2. **Handle Missing Data:** Address missing numerical values through median imputation to avoid distorting distributions.
3. **Consolidate Categorical Levels:** Group missing, blank, or generic placeholder strings (like `"Unknown Device"` or `"Invalid Method"`) into a uniform `"Unknown"` category.
4. **Feature Engineering:** Deriving a contextual `time_of_day` factor variable from the raw transaction timestamps to capture time based fraud patterns.

```{r data-cleaning, message=FALSE, warning=FALSE}
# Load required libraries
library(dplyr)

# 1. Load the raw dataset
df_raw <- read.csv("fraud_detection_dataset.csv", stringsAsFactors = FALSE)

# 2. Execute pipeline for cleaning and feature engineering
df <- df_raw %>%
  rename_with(tolower) %>%
  rename_with(~ gsub("\\.", "_", .x)) %>%
  mutate(
    # Impute missing transaction amounts with the median
    transaction_amount = ifelse(
      is.na(transaction_amount),
      median(transaction_amount, na.rm = TRUE),
      transaction_amount
    ),
    # Impute missing timestamps with the median hour
    time_of_transaction = ifelse(
      is.na(time_of_transaction),
      median(time_of_transaction, na.rm = TRUE),
      time_of_transaction
    ),
    # Standardize device categories (handles NA, placeholders, and empty strings)
    device_used = ifelse(
      is.na(device_used) | device_used %in= c("Unknown Device", "N/A", "NA", ""),
      "Unknown",
      device_used
    ),
    # Handle missing locations (fixed to catch empty strings "")
    location = ifelse(
      is.na(location) | location == "",
      "Unknown",
      location
    ),
    # Clean and standardize payment methods (fixed to catch empty strings "")
    payment_method = ifelse(
      is.na(payment_method) | payment_method %in% c("Invalid Method", ""),
      "Unknown",
      payment_method
    )
  )

# 3. Feature Engineering: Create time_of_day categorical factor
df <- df %>%
  mutate(time_of_day = case_when(
    time_of_transaction >= 6  & time_of_transaction <= 11 ~ "Morning",
    time_of_transaction >= 12 & time_of_transaction <= 17 ~ "Afternoon",
    time_of_transaction >= 18 & time_of_transaction <= 23 ~ "Evening",
    TRUE                                                  ~ "Night"
  ))

# Convert to ordered factor for logical chronological plotting
df$time_of_day <- factor(df$time_of_day,
                         levels = c("Night", "Morning", "Afternoon", "Evening"))

# Ensure target variable is numeric
df$fraudulent <- as.numeric(df$fraudulent)

# Save the cleaned dataset for modeling phases
write.csv(df, "Fraud_Detection_Dataset_Cleaned.csv", row.names = FALSE)
```

---

# Section 2: Descriptive Statistics

Before transitioning into advanced predictive modeling, it is essential to establish baseline data distributions and evaluate relationships among variables. Descriptive summaries allow us to evaluate transactional behavior across multiple categorical and continuous dimensions.

### Data Aggregation and Stratification Plan:
1. **Continuous Feature Summaries:** Generate structural summaries of transactional amounts, chronological timestamps, accounts profiles, and velocity constraints.
2. **Target Class Stratification:** Calculate detailed metric comparisons for amount variables across both fraudulent and legitimate categories.
3. **Categorical Proportional Testing:** Construct cross-tabulated contingency profiles evaluating risk ratios across hardware devices and transactional timelines.

```{r descriptive-statistics, message=FALSE, warning=FALSE}
# Load summary packages
library(dplyr)
library(tidyr)
library(psych)

# 1. Subset and summarize numerical variables
numeric_vars <- df %>%
  select(transaction_amount, time_of_transaction, previous_fraudulent_transactions,
         account_age, number_of_transactions_last_24h, fraudulent)

describe(numeric_vars)

# 2. Stratify financial values by transaction class
df %>%
  group_by(fraudulent) %>%
  summarise(
    n           = n(),
    mean_amt    = round(mean(transaction_amount), 2),
    median_amt  = round(median(transaction_amount), 2),
    sd_amt      = round(sd(transaction_amount), 2),
    min_amt     = round(min(transaction_amount), 2),
    max_amt     = round(max(transaction_amount), 2)
  )

# 3. Categorical cross-tabulation: Device types vs. Fraud status
df %>%
  group_by(device_used, fraudulent) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = fraudulent, values_from = n, values_fill = 0) %>%
  rename(Not_Fraudulent = `0`, Fraudulent = `1`) %>%
  mutate(fraud_rate = round(Fraudulent / (Not_Fraudulent + Fraudulent) * 100, 2))

# 4. Categorical cross-tabulation: Time of day vs. Fraud status
df %>%
  group_by(time_of_day, fraudulent) %>%
  summarise(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = fraudulent, values_from = n, values_fill = 0) %>%
  rename(Not_Fraudulent = `0`, Fraudulent = `1`) %>% 
  mutate(fraud_rate = round(Fraudulent / (Not_Fraudulent + Fraudulent) * 100, 2))
```

---

# Section 3: Data Visualization & Research Hypotheses

Visual analytical approaches clarify structural differences across target features. We use density plots, distributions, log scales, and fit-line overlays to systematically explore five strategic research questions regarding operational vulnerabilities.

### Research Questions Explored:
* **Exploratory Base Dist:** Review overall continuous values split by legitimate vs. anomalous target classes.
* **Research Question 1:** Evaluate variance shifts in total calculated fraud rates between unique interaction devices.
* **Research Question 2:** Map conditional probability variations of transaction risk throughout daily windows.
* **Research Question 3:** Analyze whether short-term activity velocity acts as an effective hazard predictor.
* **Research Question 4:** Contrast account maturity duration vectors against overall breach vulnerabilities.
* **Research Question 5:** Track transaction sizing dynamics to identify specific high-value spending patterns.

```{r primary-eda-plots, message=FALSE, warning=FALSE}
library(ggplot2)
library(scales)

# Set global color profiles for uniform visual storytelling
fraud_colors <- c("0" = "#4E79A7", "1" = "#E15759")

# Figure 1: Transaction Amount Density Profile
p_hist1 <- ggplot(df, aes(x = transaction_amount, fill = factor(fraudulent))) +
  geom_density(alpha = 0.5) + 
  scale_fill_manual(values = fraud_colors, labels = c("Not Fraudulent", "Fraudulent")) +
  scale_x_log10(labels = dollar_format()) + 
  labs(title = "Figure 1. Distribution of Transaction Amount",
       subtitle = "Comparing the 'shape' of spending for both groups",
       x = "Transaction Amount (USD)", y = "Density", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

# Figure 2: Account Age Density Profile
p_hist2 <- ggplot(df, aes(x = account_age, fill = factor(fraudulent))) +
  geom_density(alpha = 0.5) + 
  scale_fill_manual(values = fraud_colors, labels = c("Not Fraudulent", "Fraudulent")) +
  labs(title = "Figure 2. Distribution of Account Age",
       subtitle = "Are fraudulent accounts younger on average?",
       x = "Account Age (Days)", y = "Density", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

# Figure 3: Short-term Velocity Density Profile
p_hist3 <- ggplot(df, aes(x = number_of_transactions_last_24h, fill = factor(fraudulent))) +
  geom_density(alpha = 0.5) + 
  scale_fill_manual(values = fraud_colors, labels = c("Not Fraudulent", "Fraudulent")) +
  labs(title = "Figure 3. Distribution of Transactions in Last 24 Hours",
       subtitle = "Frequency patterns of both transaction types",
       x = "Number of Transactions (Last 24H)", y = "Density", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

# Figure 4: Historical Infraction Distribution Density
p_hist4 <- ggplot(df, aes(x = previous_fraudulent_transactions, fill = factor(fraudulent))) +
  geom_density(alpha = 0.5) + 
  scale_fill_manual(values = fraud_colors, labels = c("Not Fraudulent", "Fraudulent")) +
  labs(title = "Figure 4. Distribution of Previous Fraudulent Transactions",
       subtitle = "Historical risk profile of accounts",
       x = "Previous Fraudulent Transactions", y = "Density", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

# Figure 5: Hourly Operational Window Timeline
p_hist5 <- ggplot(df, aes(x = time_of_transaction, fill = factor(fraudulent))) +
  geom_density(alpha = 0.5) + 
  scale_fill_manual(values = fraud_colors, labels = c("Not Fraudulent", "Fraudulent")) +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  labs(title = "Figure 5. Distribution of Transaction Hour",
       subtitle = "Peak times for legitimate vs. fraudulent activity",
       x = "Hour of Day (Military Time)", y = "Density", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

# Render primary visualization assets
print(p_hist1)
print(p_hist2)
print(p_hist3)
print(p_hist4)
print(p_hist5)
```

### Research Question 1: Hardware Interface Vulnerability Analyses
We evaluate the operational breakdown across consumer devices to uncover device-specific risk concentrations within regional locations.

```{r research-question-1, message=FALSE, warning=FALSE}
# Boxplot visualization of fraud probabilities calculated by hardware interface
df %>%
  group_by(device_used, location) %>%
  summarise(fraud_rate = (sum(fraudulent) / n()) * 100, .groups = "drop") %>%
  ggplot(aes(x = reorder(device_used, -fraud_rate, FUN = mean), 
             y = fraud_rate, 
             fill = device_used)) +
  geom_boxplot(alpha = 0.85, width = 0.7) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 5, color = "black") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = percent_format(scale = 1), limits = c(0, NA)) +
  labs(title = "Figure 6. Fraud Rate Distribution by Device Type",
       subtitle = "Percentage of transactions that were fraudulent. Diamond (◆) marks the average.",
       x = "Device Used", y = "Fraud Rate (%)", fill = "Device") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

### Research Question 2: Diurnal Risk Profile Matrix
Evaluating diurnal anomalies allows us to target system monitoring efforts toward high-risk windows throughout the day.

```{r research-question-2, message=FALSE, warning=FALSE}
# 1. Quantify hourly fraud probability grouped across geographical bounds
rq2_risk_dist <- df %>%
  group_by(time_of_day, location) %>%
  summarise(
    fraud_rate_pct = (sum(fraudulent) / n()) * 100,
    .groups = "drop"
  )

# 2. Render boxplot output distributions across shifts
p_rq2_boxplot <- ggplot(rq2_risk_dist, 
                        aes(x = reorder(time_of_day, -fraud_rate_pct, FUN = mean), 
                            y = fraud_rate_pct, 
                            fill = time_of_day)) +
  geom_boxplot(alpha = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black") +
  scale_fill_brewer(palette = "Pastel1") +
  labs(title = "Figure 8. Average Fraud Probability by Time of Day",
       subtitle = "Y-axis shows the risk percentage (%). Diamond (◆) indicates the mean.",
       x = "Time of Day", y = "Probability of Fraud (%)", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

print(p_rq2_boxplot)

# 3. Print quantitative reference distributions
rq2_risk_dist %>%
  group_by(time_of_day) %>%
  summarise(
    avg_probability = round(mean(fraud_rate_pct), 2),
    min_prob = round(min(fraud_rate_pct), 2),
    max_prob = round(max(fraud_rate_pct), 2),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_probability))
```

### Research Question 3: Volume Velocity Limitations As a Fraud Predictor
We apply single-predictor generalized linear structures to evaluate the risk signals from short-term rapid activity.



```{r research-question-3, message=FALSE, warning=FALSE}
# Jitter scatter plot with embedded binomial curve overlay
p_rq3c <- ggplot(df, aes(x = number_of_transactions_last_24h, y = fraudulent)) +
  geom_jitter(height = 0.04, width = 0.15, alpha = 0.04, size = 0.8, color = "#E15759") +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
    color = "#4E79A7", linewidth = 1.2, se = TRUE) +
  scale_x_continuous(breaks = seq(floor(min(df$number_of_transactions_last_24h, na.rm = TRUE)),
                                  ceiling(max(df$number_of_transactions_last_24h, na.rm = TRUE)), by = 1)) +
  scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.25, 0.5, 0.75, 1)) +
  labs(title = "Figure 12. Logistic Regression: Fraud Probability vs. Transactions (24H)",
       subtitle = "Fitted probability curve with jittered binary fraud outcomes",
       x = "Number of Transactions (Last 24H)", y = "Probability of Fraud") +
  theme_minimal(base_size = 13)

print(p_rq3c)

# Descriptive baseline statistics
df %>%
  group_by(fraudulent) %>%
  summarise(
    n    = n(),
    mean = round(mean(number_of_transactions_last_24h), 2),
    sd   = round(sd(number_of_transactions_last_24h), 2),
    med  = median(number_of_transactions_last_24h),
    .groups = "drop"
  )

# Calculate model summary outputs
model_rq3 <- glm(fraudulent ~ number_of_transactions_last_24h, data = df, family = binomial)
summary(model_rq3)
```

### Research Question 4: Account Lifecycle Vulnerabilities
Evaluating account lifecycles helps identify potential "sleeper" accounts or immediate fraud targets.

```{r research-question-4, message=FALSE, warning=FALSE}
# Plot 4A: Stratified Outlier Boxplot Summary
p_rq4a <- ggplot(df, aes(x = factor(fraudulent), y = account_age, fill = factor(fraudulent))) +
  geom_boxplot(outlier.alpha = 0.3, outlier.size = 1) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
  scale_fill_manual(values = fraud_colors, labels = c("Not Fraudulent", "Fraudulent")) +
  scale_x_discrete(labels = c("Not Fraudulent", "Fraudulent")) +
  labs(title = "Figure 13. Account Age Distribution by Fraud Status",
       subtitle = "Diamond (◆) marks the group mean. Younger accounts show higher risk.",
       x = "Status", y = "Account Age (Days)", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

# Plot 4B: Binary Logistic Evaluation Profile
p_rq4b <- ggplot(df, aes(x = account_age, y = fraudulent)) +
  geom_point(alpha = 0.02, size = 1, color = "#E15759") + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "#4E79A7", linewidth = 1.2, se = TRUE) +
  labs(title = "Figure 14. Probability of Fraud vs. Account Age",
       subtitle = "Fitted logistic regression curve showing risk decay over time",
       x = "Account Age (Days)", y = "Probability of Fraud (0 to 1)") +
  theme_minimal(base_size = 13)

# Plot 4C: Visual Density Overlay Summary
p_rq4c <- ggplot(df, aes(x = account_age, fill = factor(fraudulent))) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = fraud_colors, labels = c("Not Fraudulent", "Fraudulent")) +
  labs(title = "Figure 15. Density of Account Ages: Fraud vs. Legitimate",
       subtitle = "Shows where the bulk of fraudulent accounts fall on the timeline",
       x = "Account Age (Days)", y = "Density", fill = "Transaction Type") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

# Render comparative group assets
print(p_rq4a)
print(p_rq4b)
print(p_rq4c)

# Statistical breakdown
df %>%
  group_by(fraudulent) %>%
  summarise(
    n    = n(),
    mean = round(mean(account_age), 2),
    sd   = round(sd(account_age), 2),
    med  = median(account_age),
    .groups = "drop"
  )

model_rq4 <- glm(fraudulent ~ account_age, data = df, family = binomial)
summary(model_rq4)
```

### Research Question 5: Value Sizing Indicators
Evaluating transaction values reveals whether anomalous activities scale proportionally with spending volume.

```{r research-question-5, message=FALSE, warning=FALSE}
# Plot 5A: Logged Amount Variant Comparison
p_rq5a <- ggplot(df, aes(x = factor(fraudulent), y = transaction_amount, fill = factor(fraudulent))) +
  geom_boxplot(outlier.alpha = 0.3, outlier.size = 1) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
  scale_fill_manual(values = fraud_colors, labels = c("Not Fraudulent", "Fraudulent")) +
  scale_x_discrete(labels = c("Not Fraudulent", "Fraudulent")) +
  scale_y_log10(labels = dollar_format()) +  
  labs(title = "Figure 16. Transaction Amount by Fraud Status (Log Scale)",
       subtitle = "Diamond (◆) marks the group mean. Log scale used to manage high outliers.",
       x = "Status", y = "Transaction Amount (USD)", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

# Plot 5B: Logistic Regression Over Looped Log Ranges
p_rq5b <- ggplot(df, aes(x = transaction_amount, y = fraudulent)) +
  geom_jitter(height = 0.05, alpha = 0.03, size = 0.8, color = "#E15759") + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "#4E79A7", linewidth = 1.2, se = TRUE) +
  scale_x_log10(labels = dollar_format()) +
  annotation_logticks(sides = "b") + 
  labs(title = "Figure 17. Probability of Fraud vs. Transaction Amount",
       subtitle = "The curve remains relatively flat even with extreme outliers.",
       x = "Transaction Amount (USD)", y = "Probability of Fraud") +
  theme_minimal(base_size = 13)

# Plot 5C: Cleaned Continuous Value Scale Density Comparison
p_rq5c <- ggplot(df, aes(x = transaction_amount, fill = factor(fraudulent))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = fraud_colors, labels = c("Not Fraudulent", "Fraudulent")) +
  scale_x_log10(labels = dollar_format()) + 
  annotation_logticks(sides = "b") +
  labs(title = "Figure 18. Density of Transaction Amounts",
       subtitle = "Log scale reveals that Fraud and Legitimate patterns are nearly identical.",
       x = "Transaction Amount (USD)", y = "Density", fill = "Transaction Type") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

# Output structural charts
print(p_rq5a)
print(p_rq5b)
print(p_rq5c)
```

---

# Section 4: Multivariate Logistic Regression Modeling

To control for confounding variables and evaluate structural hazard dependencies simultaneously, we fit a multi-predictor generalized linear model. This model isolates individual effect sizes to determine true transactional risk drivers.

### Feature Attributes Tested:
* **Financial Attributes:** `transaction_amount` (continuous asset sizes)
* **Velocity Constraints:** `number_of_transactions_last_24h` (short-term transaction counts)
* **Account Integrity Indicators:** `account_age` and historical flags (`previous_fraudulent_transactions`)
* **Contextual Profiles:** User configurations (`device_used`) and timing categories (`time_of_day`)

```{r multivariate-modeling, message=FALSE, warning=FALSE}
# 1. Fit the complete multi-predictor structural model
model_full <- glm(fraudulent ~
                    transaction_amount +
                    number_of_transactions_last_24h +
                    account_age +
                    previous_fraudulent_transactions +
                    device_used +
                    time_of_day,
                  data = df, family = binomial)

# 2. Extract and inspect estimated statistical weights
summary(model_full)

# 3. Output fit optimization summaries
cat("AIC Optimization Metric:", AIC(model_full), "\n")

# 4. Clean up environmental tracking memory
rm(list = ls())
```
